---
title: "rSwedishParliamentVotations: An API package for votations in the Swedish Parliament."
author: "Laura Julià Melis and Martin Svensson"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rSwedishParliamentVotations: An API package for votations in the Swedish Parliament}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## 1. Introduction.
The rSwedishParliamentVotations package contains a function GET_votation, used to query information for votations in the Swedish Parliment.

## 2. GET_votation function. 
GET_votation makes a GET request for data from the Swedish Parlament API, specifically for votations.

### Usage.

    GET_votation(period = NULL, span = FALSE, party = NULL, vote_result = NULL, rows = 5)
    
### Arguments.

  - **period:** a scalar or vector of year or years. The period year is defined as the year that starts of fiscal year e.g. 2018 for 2018/19. In vector form it will be e.g. [2018, 2019] will be 2018/19, 2019/20. If @param span is TRUE, then c(2017, 2019) will be evalutated as c(2017, 2018, 2019).
  - **span:** boolean argument for setting span or not for @param period.
  - **party:** string with the short names for the parties. E.g. "C" for Centerpartiet.
  - **vote_result:** string with the possible results from the voting. Possible arguments are Yes, No, Refrain and Absent.
  - **rows:** integer defining the number of results to return from the query.
 
## 3. API limitations.

This section comment some limitations in the Swedish Parliament API that has been taken into account for this package. 

#### Period.
The fiscal year range is between 2002 to present. This is due to that there is no data before this year. 

#### Party.
There are a predefined options:
```{r include=FALSE}
library(knitr)
```

```{r, results='asis', echo=FALSE}
parties.df <- data.frame("Party"= c("Centerpartiet", 
                                   "Folkpartiet", 
                                   "Liberalerna", 
                                   "Kristdemokraterna",
                                   "Miljöpartiet",
                                   "Moderata Samlingspartiet",
                                   "Socialdemokraterna",
                                   "Sverigedemokraterna",
                                   "Vänsterpartiet",
                                   "Others"), 
                        "Code"= c("C", "FP", "L", "KD", "MP", "M", "S", "SD", "V", "-" ))
kable(parties.df, align = "c")
```

Note that multiple options can be queried from the API.

#### Vote_result.

There are four diferent options: 

  - **Yes**
  - **No**
  - **Refrain**
  - **Absent**

Note that only one option can be queried at a time from the API. Otherwise, if left empty, all the options are queried.

#### Rows. 

It is not possible to query more than 10000 rows.

## 4. Example. 
This section shows how to use `GET_votation()` with some examples.

### Initialization.
```{r}
# Loading requiered packages.
library(rSwedishParliamentVotations)

# Setting the function arguments.
df <- GET_votation(period=c(2016,2018), span= TRUE, party= "C", vote_result='Ja', rows=30)
```

### Output
The function returns an object of class "data.frame" with 21 columns (variables).
```{r}
head(df)
```


## References.
* API conection: <http://data.riksdagen.se/>
* Vignettes: <http://r-pkgs.had.co.nz/vignettes.html>