---
title: 'Assignment 1: Examining multivariate data'
author: "Laura Julià Melis"
date: "11/20/2019"
output: pdf_document
---
```{r, echo = FALSE, message=FALSE}
# NEEDED LIBRARIES
library(ggplot2)
library(tidyr)
library(gridExtra)
library(car)
library(fmsb)
library(GGally)
library(RMaCzek)
```

# Question 1: Describing individual variables.
```{r, echo=FALSE}
# Importing and modifying data file
df = read.table("T1-9.dat")
colnames(df) = c("country","100m", "200m", "400m", "800m", "1500m", "3000m" , "marathon")
df1 = df[,-1]  # without the column of the countries
```

**a) Describe the 7 variables with mean values, standard deviations e.t.c.**
```{r, echo = FALSE}
# Mode: the value that appears most often
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Table with some common statistics
statsdf = data.frame(cbind(mean=sapply(df1, mean),
                       median=sapply(df1, median),
                       mode=sapply(df1, getmode),
                       min=sapply(df1, min),
                       max=sapply(df1, max),
                       range=sapply(df1, function(x)max(x)-min(x)),
                       sd=sapply(df1, sd),
                       skewness=sapply(df1, timeDate::skewness),
                       kurtosis=sapply(df1, timeDate::kurtosis)
                       ))

as.data.frame(t(round(statsdf, 4)))
```
**b) Illustrate the variables with different graphs. Are there any apparent extreme values? Do the variables seem normally distributed? Plot the best fitting Gaussian density curve on the data’s histogram.**

* Extreme values?
```{r, echo = FALSE, fig.align='center'}
# tidyr::gather
# a table with columns the type of the run and the seconds
df2 = gather(df1, "runtype", "seconds")
df22 = gather(df1[,-c(3,7)], "runtype", "seconds")  #without: 400m, marathon
df222 = gather(df1[,-c(2,3,6 ,7)], "runtype", "seconds") #without: 200m, 400m, marathon

# https://www.r-graph-gallery.com/89-box-and-scatter-plot-with-ggplot2.html
#library(gridExtra)
plot1 <- ggplot(df2, aes(x=runtype, y=seconds, fill=runtype)) + geom_boxplot()
plot2 <- ggplot(df22, aes(x=runtype, y=seconds, fill=runtype)) + geom_boxplot()
plot3 <- ggplot(df222, aes(x=runtype, y=seconds, fill=runtype)) + geom_boxplot()
grid.arrange(plot1, plot2, ncol=2)
```

In terms of boxplots (i.e. fouth quatiles) marathon has 3 outliers and is the most spread distribution. Then, 400m has one outelier and is the second most spread.


* Normally distributed?
```{r echo = FALSE, fig.align='center', fig.height=4}
#library("car")

par(mfrow=c(3,3), oma = c(0,0,0.5,0) + 0.1, mar = c(4,3,1,1) + 0.1)
for(i in 1:7){
  qqPlot(df1[,i],ylab="sample",envelope=.95,col.lines="red",pch=19,id=F) + title(main=paste(names(df1)[i], "Q-Q plot"))
}

#pch = 19: solid circle
#envelope=.95: 95% confidence level 
#id=TRUE: show the id of the outliers
```

These qq-plots will help us assess if the data is plausibly coming from a Normal distribution. The closer all points lie to the line, the closer the distribution of the sample comes to the normal distribution. Furthermore, the red line should be almost y=x (i.e. 45 degrees angle from the horizontal line).

With 95% confidence level from Quantile-Comparison Plot, we can confirm the outliers as we said from box plots before (we could even specify which observations are outliers by putting the argument id=T in qqPlot). Moreover, exept from the ouliers all variables could be normal but a bit skewed. The variables that look more normal are 100m, 200m and 400m.

In all the variables, most of the points fall in the straight line except for the points in the tails (expecially in the right tail) meaning that the variables have more extreme values than expected if they came from a Normal distribution: the empirical (or sample) quantiles are less than the theoretical quantiles.

* Gaussian density curve on the data's histogram.
```{r, echo=FALSE, fig.align='center', fig.height=3.7}
par(mfrow=c(3,3), oma = c(0,0,0.5,0) + 0.1, mar = c(4,3,1,1) + 0.1)
for(i in 1:7){
  dens = density(df1[,i])
  hist(df1[,i], freq=F, xlab="seconds",main=names(df1)[i]) #probability densities
  lines(dens,col="red")
}
```
From histograms and densities we can confirm some conclusions from before. The variables '100m', '200m' and '400m' look normal. The variables are '1500m', '3000m' and 'marathon' look skewed and from the table of 1.a we can confirm that their skewed is high; 1.943142,	2.5372744,	2.311548 respectively.

# Question 2: Relationships between the variables.

**a) Compute the covariance and correlation matrices for the 7 variables.**

* Covariance matrix:
```{r, echo=FALSE}
(S <- round(cov(df1),4))
```
All the covariances between variables are positive, meaning that all pair of variables are positively related. Also, we can observe that the covariances are smaller between those variables representing short races and larger between variables that represent linger races (such as "Marathon"). The highest covariance coefficient is $\text{Cov}(\text{400m},\text{Marathon}) = 28.9$ and the lowest, $\text{Cov}(\text{800m},\text{1500m}) = 0.0214$.

* Correlation matrix:
```{r, echo=FALSE}
(R <- round(cor(df1),4))
```
The correlation coefficient is a measure that calculates the strength and direction of the linear relationship between two variables ($r \in \{-1,+1\}$). We observe that all the coefficients are positive (and greater than $0.65$) so all pairs of variables have a positive linear relationship meaning that, as the value of one variable increases, the value of the other variable increases too. The highest coefficient is $\text{Cor}(\text{1500m},\text{3000m}) = 0.9734$ and the lowest, $\text{Cor}(\text{100m},\text{Marathon}) = 0.6690$.


**b) Generate and study the scatterplots between each pair of variables. Any extreme values?**
```{r, echo=FALSE, message=FALSE, fig.align='center', fig.height=4.1, fig.width=8}
#library(GGally)
ggpairs(df1,progress=F) 
```
From pair scatterplot, we could see outliers at almost every pair (e.g. '200m' vs '800m','400m' vs '800m', '400m' vs '1500m'), but is is highly seen on every valiable vs 'marathon'.

**c) Present other (at least two) graphs that you find interesting with respect to this data set.**

* Scatterplots with labeled extreme points:

In the following plots, green lines represent the mean and red lines represent the lower and upper limit. These limits have been calculated as follows:

$$\bar{x} \pm 2\sigma \text{ because  Pr}\{\mu-2\sigma \le X \le\mu+2\sigma\}=0.9545$$

```{r echo = FALSE}
par(mfrow=c(3,3), oma = c(0,0,0.5,0) + 0.1, mar = c(4,3,1,1) + 0.1)
for(i in 2:8){
 low_dist <- mean(df[,i])-2*sd(df[,i])  # 95.45% of the values are inside this interval:
 upp_dist <- mean(df[,i])+2*sd(df[,i])
 extreme <- which(df[,i] > upp_dist | df[,i] < low_dist)

plot(df[,i], df[,1], xlab="Country", ylab=names(df[i]), main=names(df)[i])+
  abline(v=c(low_dist,upp_dist, mean(df[,i])), col=c("red","red","green")) + 
  text(df[extreme,i], df[extreme,1],as.vector(df[extreme,1]))
}
```
So, looking at these plots, we can observe that the most extreme countryes are COK, SAM and PNG.

* 3D Scatterplots:

We will take the 3 most uncorrelated variables, in order to have the most information, to plot in 3D scatterplot.

```{r echo = FALSE, fig.align='center'}
df3 = df[,c(2,6,8)]
rownames(df3) = df$country
stars(df3, full = T, key.loc = c(20, 2))  #"100m", "1500m" and "marathon"

d = df[,-1]
rownames(d) = df$country
stars(d, full = T, key.loc = c(20, 2))
```

* Radar charts:
```{r echo = FALSE}
# #library(fmsb)
# #radar chart
# min_row <- sapply(df, min)
# max_row <- sapply(df, max)
# data_radar <- rbind(min_row, max_row, df)
# data_radar[1,1] <- "min"
# data_radar[2,1] <- "max"
# data_radar[,2:8] <- lapply(data_radar[,2:8], function(x) as.numeric(as.character(x)))
# 
# # countries c(1,2, ..)
# radarchart(data_radar[1:5,2:8])
```

# Question 3: Examining for extreme values.

**a) Look at the plots (esp. scatterplots) generated in the previous question. Which 3–4 countries appear most extreme? Why do you consider them extreme?One approach to measuring “extremism” is to look at the distance (needs to be defined!) between an observation and the sample mean vector, i.e. we look how far one is from the average. Such a distance can be called an multivariate residual for the given observation.**

As we can see from star plots above, the most extreme countries in terms of good performance are United States (USA), Russia (RU), China (CHN) and Germany (GER). In terms of bad performance the most extreme countries are Samoa (SAM), Cook Islands (COK), Papua New Guinea (PNG) and Guatemala (GUA).

We consider them extreme because the values of the times for the runtypes are either too small (good performance) or too big (bad performance).

**b) The most common residual is the Euclidean distance between the observation and sample mean vector, i.e.**
$$d(\vec{x},\bar{x})= \sqrt{(\vec{x}-\bar{x})^T\cdot (\vec{x}-\bar{x})}$$
**This distance can be immediately generalized to the $L^r, r > 0$ distance as **
$$d_{L^r}(\vec{x},\bar{x})=\Bigg( \sum^p_{i=1}|\vec{x}_i-\bar{x}_i|^r\Bigg)^{1/r}$$
**where p is the dimension of the observation (here $p = 7$).**

**Compute the squared Euclidean distance (i.e. $r = 2$) of the observation from the sample mean for all 55 countries using R’s matrix operations. First center the raw data by the means to get $\vec{x}-\bar{x}$ for each country. Then do a calculation with matrices that will result in a matrix that has on its diagonal the requested squared distance for each country. Copy this diagonal to a vector and report on the five most extreme countries. In this questions you MAY NOT use any loops.**

```{r echo = FALSE}
resid = sapply(df[,-1],function(x)x-mean(x))  #center the raw data by the means
euclidist = sqrt(abs(resid%*%t(resid)))  #Euclidean distance between the observation and sample mean vector

dists = diag(euclidist)   #diag is squared distance for each country
names(dists) = df$country 
head(sort(dists, decreasing=TRUE),5)  #sort with decreasing order and take the first five
```

The five most extreme countries are Papua New Guinea (PNG), Cook Islands (COK), Samoa (SAM), Bermuda (BER) and United Kingdom (GBR). 


**c) The different variables have different scales so it is possible that the distances can be domi- nated by some few variables. To avoid this we can use the squared distance**
$$d_{\boldsymbol{V}}^2(\vec{x},\bar{x})=(\vec{x}-\bar{x})^T\boldsymbol{V}^{-1}(\vec{x}-\bar{x}),$$

**where $\boldsymbol{V}$ is a diagonal matrix with variances of the appropriate variables on the diagonal. The effect, is that for each variable the squared distance is divided by its variance and we have a scaled independent distance.**

**It is simple to compute this measure by standardizing the raw data with both means (cen- tring) and standard deviations (scaling), and then compute the Euclidean distance for the normalized data. Carry out these computations and conclude which countries are the most extreme ones. How do your conclusions compare with the unnormalized ones?**

```{r echo = FALSE}
V = diag(diag(S)) #first diag to take the diagonal 
                  #then diag again to make the vector a diagonal matrix
sqdist = resid%*%solve(V)%*%t(resid)

sqdist = diag(sqdist)
names(sqdist) = df$country 

head(sort(sqdist, decreasing=TRUE),5)  #sort with decreasing order and take the first five
```

Uding the Euclidean distance for thenormalized data, the five most extreme countries are Samoa (SAM), Cook Islands (COK), Papua New Guinea (PNG), United States (USA) and Singapore (SIN). 

**d) The most common statistical distance is the Mahalanobis distance **
$$d_M^2(\vec{x},\bar{x})=(\vec{x}-\bar{x})^T\boldsymbol{C}^{-1}(\vec{x}-\bar{x}),$$

**where $\boldsymbol{C}$ is the sample covariance matrix calculated from the data. With this measure we also use the relationships (covariances) between the variables (and not only the marginal variances as $d_{\boldsymbol{V}}(·,·)$ does). Compute the Mahalanobis distance, which countries are most extreme now?**

```{r echo = FALSE}
mahdist = resid%*%solve(S)%*%t(resid)

mahdist = diag(mahdist)
names(mahdist) = df$country 

head(sort(mahdist, decreasing=TRUE),5)  #sort with decreasing order and take the first five
```

Using the Mahalanobis distance, the five most extreme countries are Samoa (SAM), Papua New Guinea (PNG), North Korea (KORN), Cook Islands (COK) and Mexico (MEX). 

**e) Compare the results in b)–d). Some of the countries are in the upper end with all the measures and perhaps they can be classified as extreme. Discuss this. But also notice the different measures give rather different results (how does Sweden behave?). Summarize this graphically. Produce Czekanowski’s diagram using e.g. the `RMaCzek` package. In case of problems please describe them.**

1. COMPARISON: 

         Results in b): PNG > COK > SAM > BER > GBR
         Results in d): SAM > PNG > KORN > COK > MEX
    
In b, where the Eucliden distance (using the centered data by its mean) has been used, we haven't considered the variances and as a result, those variables with a diferent scale has dominated the distances: PNG, COK and SAM are three countries with extreme values in those variables where the times are scaled in minutes. 

On the other hand, when we have calculated the Mahalanobis distance, we have divided each variable by its variance and also have considered the covariance between each pair of variables. As a result, we can see that for example the country KORN (that has not appeared in the plots analyzed in the questions above) is the third most extreme country. 

Also, we can see that in b) SAM has obtained a distance of 38.52476 (third position), with a difference of 29 points with the first one while in d) has been the most extreme country. This might be explained because PNG and COK are quite more extreme in the marathon variable than SAM and this variable is dominating a lot the distance between the observations and the mean.

2. DISCUSSION:

In all the measures, SAM, PNG and COK have been the ones with biggest distances. This makes sense because in all the plots we have seen that these 3 countries have been always the ones with highest times for all the variables. 

3. SWEEDEN:

In the next plot, each point represents a country (blue points represent Sweden), x-axis shows he three different distance measures that we have used and y-axis indicates the distance value.

(plot)

We observe that sweeden is in the three cases quite close to the mean (distance is close to 0)...

4. CZEKANOWSKI'S DIAGRAM:
Czekanowski's function calculates the distance (by default, the Euclidean distance) between all possible pairs of objects (in our case, countries). 
```{r echo = FALSE, fig.align='center', fig.height=8}
#library(RMaCzek)
df3d = df1
rownames(df3d) = df$country 
m = czek_matrix(df3d)
plot.czek_matrix(m)
```

From the diagram we can conclude that USA, CHN and RUS (top left corner) are the three countries with the greatest lower distances (countries that performe better, they are far from the mean values in all variables) while again, in the bottom right corner, we see that PNG, COK and SAM are the three most extreme values, with the greatest upper distances (worst performances).

# Appendix.
## Question 1

```{r, eval=FALSE}
# NEEDED LIBRARIES
library(ggplot2)
library(tidyr)
library(gridExtra)
library(car)
library(fmsb)
library(GGally)
library(RMaCzek)

# Importing and modifying data file
df = read.table("T1-9.dat")
colnames(df) = c("country","100m", "200m", "400m", "800m", "1500m", "3000m" , "marathon")
df1 = df[,-1]  # without the column of the countries
```

**Subquestion a.**
```{r, eval=FALSE}
# Mode: the value that appears most often
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Table with some common statistics
statsdf = data.frame(cbind(mean=sapply(df1, mean),
                       median=sapply(df1, median),
                       mode=sapply(df1, getmode),
                       min=sapply(df1, min),
                       max=sapply(df1, max),
                       range=sapply(df1, function(x)max(x)-min(x)),
                       sd=sapply(df1, sd),
                       skewness=sapply(df1, timeDate::skewness),
                       kurtosis=sapply(df1, timeDate::kurtosis)
                       ))

as.data.frame(t(round(statsdf, 4)))
```

**Subquestion b.**
```{r, eval=FALSE}
# b1. EXTREME VALUES?
# tidyr::gather
# a table with columns the type of the run and the seconds
df2 = gather(df1, "runtype", "seconds")
df22 = gather(df1[,-c(3,7)], "runtype", "seconds")  #without: 400m, marathon
df222 = gather(df1[,-c(2,3,6 ,7)], "runtype", "seconds") #without: 200m, 400m, marathon

# https://www.r-graph-gallery.com/89-box-and-scatter-plot-with-ggplot2.html
#library(gridExtra)
par(mfrow=c(1,3))
ggplot(df2, aes(x=runtype, y=seconds, fill=runtype)) + geom_boxplot()
ggplot(df22, aes(x=runtype, y=seconds, fill=runtype)) + geom_boxplot()
ggplot(df222, aes(x=runtype, y=seconds, fill=runtype)) + geom_boxplot()

# b2. NORMALLY DUSRIBUTED?
#library("car")
par(mfrow=c(3,3), oma = c(0,0,0.5,0) + 0.1, mar = c(4,3,1,1) + 0.1)
for(i in 1:7){
  qqPlot(df1[,i],ylab="sample",envelope=.95,col.lines="red",pch=19,id=F) + title(main=paste(names(df1)[i], "Q-Q plot"))
}

#pch = 19: solid circle
#envelope=.95: 95% confidence level 
#id=TRUE: show the id of the outliers

# b3. GAUSSIAN DENSITY CURVE. 
par(mfrow=c(3,3), oma = c(0,0,0.5,0) + 0.1, mar = c(4,3,1,1) + 0.1)
for(i in 1:7){
  dens = density(df1[,i])
  hist(df1[,i], freq=F, xlab="seconds",main=names(df1)[i]) #probability densities
  lines(dens,col="red")
}
```

## Question 2
**Subquestion a.**
```{r, eval=FALSE}
(S <- round(cov(df1),4)) # Covariance matrix
(R <- round(cor(df1),4)) # Correlation matrix
```
**Subquestion b.**
```{r, eval=FALSE}
#library(GGally)
ggpairs(df1,progress=F) 
```
**Subquestion c.**
```{r, eval=FALSE}
# C1. Scatterplots with labelled extreme points
par(mfrow=c(3,3), oma = c(0,0,0.5,0) + 0.1, mar = c(4,3,1,1) + 0.1)
for(i in 2:8){
 low_dist <- mean(df[,i])-2*sd(df[,i])  # 95.45% of the values are inside this interval:
 upp_dist <- mean(df[,i])+2*sd(df[,i])
 extreme <- which(df[,i] > upp_dist | df[,i] < low_dist)

plot(df[,i], df[,1], xlab="Country", ylab=names(df[i]), main=names(df)[i])+
  abline(v=c(low_dist,upp_dist, mean(df[,i])), col=c("red","red","green")) + 
  text(df[extreme,i], df[extreme,1],as.vector(df[extreme,1]))
}

# C2. 3d scatterplots
df3 = df[,c(2,6,8)]
rownames(df3) = df$country
stars(df3, full = T, key.loc = c(20, 2))  #"100m", "1500m" and "marathon"

d = df[,-1]
rownames(d) = df$country
stars(d, full = T, key.loc = c(20, 2))

# C3. Radar chart
# #library(fmsb)
# #radar chart
# min_row <- sapply(df, min)
# max_row <- sapply(df, max)
# data_radar <- rbind(min_row, max_row, df)
# data_radar[1,1] <- "min"
# data_radar[2,1] <- "max"
# data_radar[,2:8] <- lapply(data_radar[,2:8], function(x) as.numeric(as.character(x)))
# 
# # countries c(1,2, ..)
# radarchart(data_radar[1:5,2:8])
```

## Question 3

**Subquestion b.**
```{r, eval=FALSE}
resid = sapply(df[,-1],function(x)x-mean(x))  #center the raw data by the means
euclidist = sqrt(abs(resid%*%t(resid)))  #Euclidean distance between the observation and sample mean vector

dists = diag(euclidist)   #diag is squared distance for each country
names(dists) = df$country 
head(sort(dists, decreasing=TRUE),5)  #sort with decreasing order and take the first five
```

**Subquestion c.**
```{r eval = FALSE}
V = diag(diag(S)) #first diag to take the diagonal 
                  #then diag again to make the vector a diagonal matrix
sqdist = resid%*%solve(V)%*%t(resid)

sqdist = diag(sqdist)
names(sqdist) = df$country 

head(sort(sqdist, decreasing=TRUE),5)  #sort with decreasing order and take the first five
```
**Subquestion d.**
```{r eval = FALSE}
mahdist = resid%*%solve(S)%*%t(resid)

mahdist = diag(mahdist)
names(mahdist) = df$country 

head(sort(mahdist, decreasing=TRUE),5)  #sort with decreasing order and take the first five
```

**Subquestion e.**
```{r eval = FALSE}
#library(RMaCzek)
df3d = df1
rownames(df3d) = df$country 
m = czek_matrix(df3d)
plot.czek_matrix(m)
```
